### 2021-06-21 paper.tex
introduction.

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\numberwithin{equation}{section}
\usepackage{slashed}
\usepackage{braket}
\usepackage[svgnames,psnames]{xcolor}
\usepackage[colorlinks,citecolor=DarkGreen,linkcolor=FireBrick,urlcolor=FireBrick,linktocpage,unicode]{hyperref}
\urlstyle{rm}
%\usepackage[colorlinks,citecolor=black,linkcolor=black]{hyperref}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{tikz}
%\usepackage{tikz-cd}
\tikzset{>=stealth}

@@

\def\SU{\mathrm{SU}}
\def\SO{\mathrm{SO}}
\def\so{\mathfrak{so}}
\def\Spin{\mathrm{Spin}}
\def\boo{0.0}
\def\xlattice#1#2#3{
\begin{tikzpicture}[scale=.5]
\filldraw[color=black!5!white](-.5,-.5) rectangle (1.5,1.5);
\draw[->] (-1,0) -- (2,0);
\draw[->] (0,-1) -- (0,2);
\foreach \x in {0,1} {
	\foreach \y in {0,1}{
		\pgfmathsetmacro\a{mod(#1 * \x - #2 * \y,2)}
		\ifx\a\boo
			\filldraw[color=#3] (\x,\y) circle (.5em);
		\else
			\filldraw[fill=white,draw=gray] (\x,\y) circle (.5em);
		\fi
	}
}
\end{tikzpicture}
}

@@

{\Large \bfseries Higher symmetries and anomalies \\[1em]
in $\mathfrak{so}$ QCD  and \Nequals1 duality}

@@

\noindent
We study higher symmetries and anomalies of 4d  $\mathfrak{so}(2n_c)$ gauge theory with $N_f=2n_f$ flavors.
We find that they depend on the parity of $n_c$ and $n_f$,
on the global form of the gauge group, and the discrete theta angle.
The contribution from the fermions plays a central role in our analysis.
Furthermore, our conclusion applies to \Nequals1 supersymmetric cases as well, and
we see that higher symmetries and anomalies match across the duality 
$\mathfrak{so}(2n_c)\leftrightarrow\mathfrak{so}(2n_f-n_c+4)$ of Intriligator and Seiberg.

@@

As for the 0-form symmetry, 
we focus our attention on the $\SU(N_f)$ symmetry acting on $N_f$ flavors of matter fields 
in the vector representation.
There can be and definitely are other discrete symmetries, but we will not consider them in this paper for brevity.

Our main question is then how the $\bZ_2$ 1-form symmetry and the $\SU(N_f)$ 0-form symmetry are related.\footnote{%
A partial answer was given in \cite{Hsin:2020nts}, but the contribution from fermions was not taken into account in that reference.
}
We concentrate on the case when $N_c$ and $N_f$ are both even, $N_c=2n_c$ and $N_f=2n_f$.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\numberwithin{equation}{section}
\usepackage{slashed}
\usepackage{braket}
\usepackage[svgnames,psnames]{xcolor}
\usepackage[colorlinks,citecolor=DarkGreen,linkcolor=FireBrick,urlcolor=FireBrick,linktocpage,unicode]{hyperref}
\urlstyle{rm}
%\usepackage[colorlinks,citecolor=black,linkcolor=black]{hyperref}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{tikz}
%\usepackage{tikz-cd}
\newcommand{\tikzmark}[1]{\tikz[remember picture,overlay]\node (#1){};}
\tikzset{>=stealth}

@@

\def\SU{\mathrm{SU}}
\def\SO{\mathrm{SO}}
\def\su{\mathfrak{su}}
\def\so{\mathfrak{so}}
\def\Spin{\mathrm{Spin}}
\def\boo{0.0}
\def\xlattice#1#2#3{
\begin{tikzpicture}[scale=.5]
\filldraw[color=black!5!white](-.5,-.5) rectangle (1.5,1.5);
\draw[->] (-1,0) -- (2,0);
\draw[->] (0,-1) -- (0,2);
\foreach \x in {0,1} {
	\foreach \y in {0,1}{
		\pgfmathsetmacro\a{mod(#1 * \x - #2 * \y,2)}
		\ifx\a\boo
			\filldraw[color=#3] (\x,\y) circle (.5em);
		\else
			\filldraw[fill=white,draw=gray] (\x,\y) circle (.5em);
		\fi
	}
}
\end{tikzpicture}
}

@@

{\Large \bfseries Higher symmetries and anomalies \\[1em]
in $\so$ QCD  and \Nequals1 duality}

@@

\noindent
We study higher symmetries and anomalies of 4d  $\so(2n_c)$ gauge theory with $N_f=2n_f$ flavors.
We find that they depend on the parity of $n_c$ and $n_f$,
on the global form of the gauge group, and the discrete theta angle.
The contribution from the fermions plays a central role in our analysis.
Furthermore, our conclusion applies to \Nequals1 supersymmetric cases as well, and
we see that higher symmetries and anomalies match across the duality 
$\so(2n_c)\leftrightarrow\so(2n_f-n_c+4)$ of Intriligator and Seiberg.

@@

As for the 0-form symmetry, 
we focus our attention on the $\su(N_f)$ symmetry acting on $N_f$ flavors of matter fields 
in the vector representation.
There can be and definitely are other discrete symmetries, but we will not consider them in this paper for brevity.

The main question is then how the $\bZ_2$ 1-form symmetry and the $\su(N_f)$ 0-form symmetry are related.\footnote{%
A partial answer was given in \cite{Hsin:2020nts}, but the contribution from fermions was not taken into account in that reference.
Our conclusion is therefore somewhat different from theirs.
}
We concentrate on the case when $N_c$ and $N_f$ are both even, $N_c=2n_c$ and $N_f=2n_f$.

Take for example the $Spin(2n_c)$ gauge theory with $2n_f$ flavors,
when $n_c$ is odd.
Take two copies of  the Wilson line $W$ in the spinor representation. 
They form a Wilson line in the vector representation.
This can be screened by a dynamical fermion, which was why $W^2=1$ 
as far as the 1-form symmetry charge was concerned.
Now let us recall that this dynamical fermion transforms nontrivially under $-1\in \SU(2n_f)$.
Therefore, when we take the flavor symmetry into account, $W^2$ is still nontrivial.
As we will recall below,  formally this means that the $\bZ_2$ 1-form symmetry extends the $\SU(2n_f)/\bZ_2$ 0-form symmetry in a nontrivial manner, forming a nontrivial 2-group  $H$\begin{equation}
0\to \bZ_2[1] \to H\to \SU(2n_f)/\bZ_2\to 0,\label{2-group}
\end{equation}
whose Postnikov class is specified by \begin{equation}
\beta v_2 \in H^3(\SU(2n_f)/\bZ_2,\bZ_2),\label{postnikov}
\end{equation}
where $a_2$ is the generator of $H^2(\SU(2n_f)/\bZ_2,\bZ_2)=\bZ_2$ and $\beta$ is the Bockstein.

The $\SO_+$ gauge theory is then obtained by gauging the $\bZ_2$ 1-form symmetry \cite{Kapustin:2014gua}.
The presence of the fermions significantly complicates the analysis.
For the moment let us suppose that we have $N_f$ scalars instead of fermions in the vector representation.
Then, the argument of \cite{Tachikawa:2017gyf} immediately applies, and 
we see that the $\bZ_2$ 1-form symmetry of the $\SO(2n_f)$ theory and the $\SU(2n_f)/\bZ_2$ 0-form flavor symmetry remains a direct product but with a mixed anomaly given by \begin{equation}
2\pi i \frac12 \int B\beta v_2.\label{mixed}
\end{equation}

In the rest of the paper, we will carefully analyze how the $\bZ_2$ 1-form symmetry and the $\SU(2n_f)/\bZ_2$ 0-form symmetry are combined.
The derivation will be detailed in the following, and here we simply summarize the result  in Table~\ref{table:main}.
There, `none' specifies that they remain a direct product without mixed anomaly;
`anomaly' implies that they remain a direct product but with mixed anomaly of the form \eqref{mixed};
and `extension' means that they combine into a 2-group given by \eqref{2-group} with the Postnikov class \eqref{postnikov}.

\begin{table}
\centering
\begin{tabular}{r@{\vphantom{$\Bigm|$}\,}l|ccc}
$(n_c,$&$n_f)$ & $\Spin$ & $\SO_+$ & $\SO_-$\\
\hline
(even,&even) & no\tikzmark{A}ne & no\tikzmark{B}ne & no\tikzmark{C}ne \\
(odd,&even) & exte\tikzmark{P}nsion & ano\tikzmark{Q}maly & exte\tikzmark{R}nsion \\
(even,&odd) & ano\tikzmark{S}maly & exte\tikzmark{T}nsion & exte\tikzmark{U}nsion \\
(odd,&odd) & exte\tikzmark{L}nsion & exte\tikzmark{M}nsion & ano\tikzmark{N}maly 
\end{tabular}
\caption{How the $\bZ_2$ 1-form symmetry and the $\SU(2n_f)/\bZ_2$ 0-form symmetry are combined
in $\so(2n_c)$ QCD.
`none' implies that they remain a direct product without mixed anomaly;
`anomaly' means that they remain a direct product but with mixed anomaly;
and `extension' is when they combine into a nontrivial 2-group. 
The orange lines show how the duality of Intriligator and Seiberg acts on this set of theories.
\label{table:main}}
\end{table}
\tikz[overlay,remember picture]{%
\draw[<->,bend left,color=Orange,line width=1.5] (A.north) to (C.north);
\draw[<->,color=Orange,line width=1.5] (B.north west) .. controls +(80:0.5) ..  (B.north east);
\draw[<->,bend left,color=Orange,line width=1.5] (P.north) to (R.north);
\draw[<->,color=Orange,line width=1.5] (Q.north west) .. controls +(80:0.5) ..  (Q.north east);
\draw[<->,color=Orange,line width=1.5] (S.south) to (N.north);
\draw[<->,color=Orange,line width=1.5] (T.south) to (M.north);
\draw[<->,color=Orange,line width=1.5] (U.south) to (L.north);
}

Our result is equally applicable in the case of \Nequals1 supersymmetric QCD, for which
Intriligator and Seiberg found in \cite{Intriligator:1995id} a duality exchanging $\so(N_c)$ and $\so(N_f-N_c+4)$,
which in our notation sends $n_c$ to $n_c'=n_f-n_c+2$.
In \cite{Aharony:2013hda}, this duality was refined to account for the global form of the gauge group and the discrete theta angle, and it was concluded that $\Spin$ is exchanged with $\SO_-$ while $\SO_+$ maps to itself.
This mapping was checked using supersymmetric localization on $S^3/\bZ_n \times S^1$ in \cite{Razamat:2013opa}.
Our analysis allows us to check this duality by comparing how the 1-form symmetry and the 0-form symmetry are combined in the dual pairs.
We superimposed the action of the duality on our main Table~\ref{table:main}.
It is satisfying to see that the duality action correctly preserves the labels `none', `anomaly' and `extension'.

The rest of the paper is organized as follows ...
